---
title: "Programming with models: Inspecting model structure"
subtitle: "ENVR Asheville workshop short course"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found 
library(nimble)
read_chunk("chunks_litters.R")
```

# Running example: the AFT model

Now let's explore how to understand the dependence structure of a model. After all the model is a directed acyclic graph, so we should be able to work with that graph structure.

We'll demo this using the AFT survival model from the [example models](example_models_slides.Rmd). I'll make a slight tweak to the model and use a multivariate normal prior distribution for *delta* in order to illustrate some additional concepts.

We first need to build the model. I'll build a version using a different parameterization for the survival probabilities.

```{r, litters-reparam-code}
littersCode <- nimbleCode({
  for (i in 1:G) {
     for (j in 1:N) {
        # likelihood (data model)
        r[i,j] ~ dbin(p[i,j], n[i,j])
        # latent process (random effects)
        p[i,j] ~ dbeta(mean = mu[i], sd = sigma[i])
     }
     # prior for hyperparameters
     mu[i] ~ dflat()
     sigma[i] ~ dunif(0, 100)
   }
})
```

```{r, aft-setup}
G <- 2
N <- 16
n <- matrix(c(13, 12, 12, 11, 9, 10, 
              9, 9, 8, 11, 8, 10, 13, 10, 12, 9, 10, 9, 10, 5, 9, 9, 13, 
              7, 5, 10, 7, 6, 10, 10, 10, 7), nrow = 2)
r <- matrix(c(13, 12, 12, 11, 9, 10, 9, 9, 8, 10, 8, 9, 
     12, 9, 11, 8, 9, 8, 9, 4, 8, 7, 11, 4, 4, 5, 5, 3, 7, 3, 7, 0), 
     nrow = 2)
              
littersConstants <- list(G = G, N = N, n = n)
littersData <- list(r = r)
littersInits <- list(mu = rep(0.5, 2), sigma = rep(3, 2))

## create the NIMBLE model object
model <- nimbleModel(littersCode, 
          data = littersData, constants = littersConstants, inits = littersInits)
```


# Nodes and variables

The parameters and data in a model are represented as nodes in a graph. Here we define some terms:

 - parameter: an unknown quantity in the model that is represented as a random variable and will generally be estimated in an algorithm, e.g. `mu[1]` in the litters model
 - data: a known quantity in a model, also represented as a random variable, e.g., `r[1, 1], ..., r[1, 16], ...` in the litters model
 - constants: other fixed quantities involved in the model, e.g., `G` in the AFT model
 - node: an element in the model graph representing data, parameter, or a deterministic quantity that is a function of other quantities in the model, e.g., `mu[1]`, `mu[2]`, or `p[1, 5]` in the AFT model
 - variable: a collection of one or more nodes with the same name, e.g., `mu` or `r` in the litters model. 

Note, if we have a multivariate distribution, then the node includes all of the elements: suppose we define `mu[1:2]` to have a bivariate normal prior; then `mu[1:2]` is the node and `mu` is the variable. 

# Querying nodes and variables

NIMBLE provides users and programmers with the ability to get information about the nodes and variables in the model and the relationships amongst them.

What are the nodes in the model? What are the dependencies of a given node (this is information used in many algorithms)?


```{r, model-nodes}
nodes <- model$getNodeNames()
nodes[1:11]
top <- model$getNodeNames(topOnly = TRUE)
top
muDeps <- model$getDependencies('mu')
muDeps[1:11]
```

# Data nodes

```{r, model-data}
model$getNodeNames(dataOnly = TRUE)
model$isData('r')
## parameters (including imputed data)
model$getNodeNames(stochOnly = TRUE, includeData = FALSE)
```

# More details on querying model structure

There are a variety of options to `getNodeNames()` and `getDependencies` that allow you to fine-tune the information you get.

```{r, get-info}
args(model$getDependencies)
args(model$getNodeNames)

latents <- model$getNodeNames(latentOnly = TRUE, stochOnly = TRUE, includeData = FALSE)
latents
model$getDependencies(latents, dataOnly = TRUE)
model$getNodeNames(dataOnly = TRUE)
```

# Inspecting the model: variables

What are the variables in the model? What information is available about them?

```{r, model-vars}
model$getVarNames()
model$getVarInfo('mu')
```

As of the current version of NIMBLE, information about
variables is not competely nicely arranged for a user (there aren't as
many query functions), but it is available.

This variable has 1 dimension (`nDim`), and its size is 2 (`maxs`).  Currently `mins` is always 1.  If at least one node within the variable is stochastic, then `anyStoch` will be `TRUE`.


# The importance of querying a model: lifted nodes

You may have noticed a few strangely-named nodes, such as "lifted_mu_oBi_cB_tothe_2_times__oP1_minus_mu_oBi_cB_cP_over_sigma_oBi_cB_tothe_2_minus_mu_oBi_cB_L4[1]", in the model. NIMBLE implements some features by inserting its own nodes. In this case we reparameterize the model from the mean/sd parameterization of a beta distribution to the shape1/shape2 parameterization, and "lifted_mu_oBi_cB_tothe_2_times__oP1_minus_mu_oBi_cB_cP_over_sigma_oBi_cB_tothe_2_minus_mu_oBi_cB_L4[1]" is the first shape parameter of the beta distribution for the first group.

*You should never assume you know what nodes are in a model simply
because you wrote the BUGS code.*

Let's look at the two main ways this happens:

```{r, lifted-1, fig.cap=""}
m1 <- nimbleModel(
    nimbleCode({
        tau ~ dunif(0, 100)
        x ~ dnorm(0, tau) #by default, tau is a precision
    }))
plot(m1$getGraph())
m1$getNodeNames()
```
The node `lifted_d1_over_sqrt_oPtau_cP` has been inserted between
`tau` and `x`.   The resulting model would equivalently have been
created by this BUGS code:

```{r, lifted-2, eval = FALSE}
nimbleCode({
    tau ~ dunif(0, 100)
    lifted_d1_over_sqrt_oPtau_cP <- 1/sqrt(tau)
    x ~ dnorm(0, sd = lifted_d1_over_sqrt_oPtau_cP) # override and make 2nd arg the SD
})
```

NIMBLE has *lifted* the calculation of standard deviation from
precision so that it is part of the model's graph.  Therefore *you
will make a mistake if you assume that the dependencies of `tau`
include only `x`*:

```{r, bad}
m1$tau <- 3
m1$x <- 1
m1$calculate(c('tau','x')) ## Wrong: the lifted node is being neglected
```

The correct way to do this is with model-generic programming:
```{r, model-generic}
m1$getDependencies('tau')
m1$calculate( m1$getDependencies('tau') )
```

